Inference of proto-neutron star properties from gravitational-wave data in core-collapse supernovae

### 2020-11-08 ccsn.tex
Some work on the intro.
Some further work on the paper.
More work. Almost done.
Finished my revision of the entire paper.

\definecolor{maob}{rgb}{0.6,0.1,0.2}
\newcommand{\maob}[1]{\textcolor{maob}{#1}}
\newcommand{\pcd}[1]{\textcolor{blue}{#1}}
\newcommand{\mab}[1]{\textcolor{red}{#1}}
\newcommand{\tf}[1]{\textcolor{green}{#1}}

@@

\begin{document}
\linenumbers

\title{Inference of proto-neutron star properties from gravitational-wave data in core-collapse supernovae.}

\author{M.~A.~Bizouard}
\affiliation{Artemis, Universit\'e C\^ote d'Azur, Observatoire C\^ote d'Azur, CNRS, CS 34229, F-06304 Nice Cedex 4, France}

\author{P.~Maturana-Russel}
\affiliation{Department of Statistics, The University of Auckland, Auckland, New Zealand}
\affiliation{Department of Mathematical Sciences, Auckland University of Technology, Auckland, New Zealand}

\author{A.~Torres-Forn\'e}
\affiliation{Max Planck Institute for Gravitationalphysik (Albert Einstein Institute), D-14476 Potsdam-Golm, Germany}
\affiliation{Departamento de Astronom\'{\i }a y Astrof\'{\i }sica, Universitat de Val\`encia, E-46100 Burjassot, Val\`encia, Spain}

\author{M.~Obergaulinger}
\affiliation{Departamento de Astronom\'{\i }a y Astrof\'{\i }sica, Universitat de Val\`encia, E-46100 Burjassot, Val\`encia, Spain}

\author{P.~Cerd\'a-Dur\'an}
\affiliation{Departamento de Astronom\'{\i }a y Astrof\'{\i }sica, Universitat de Val\`encia, E-46100 Burjassot, Val\`encia, Spain}

\author{N.~Christensen}
\affiliation{Artemis, Universit\'e C\^ote d'Azur, Observatoire C\^ote d'Azur, CNRS, CS 34229, F-06304 Nice Cedex 4, France}
\affiliation{Carleton College, Northfield, MN 55057, USA}

\author{J.~A.~Font}
\affiliation{Departamento de Astronom\'{\i }a y Astrof\'{\i }sica, Universitat de Val\`encia, E-46100 Burjassot, Val\`encia, Spain}
\affiliation{Observatori Astron\`omic, Universitat de Val\`encia, E-46980, Paterna, Val\`encia, Spain}

\author{R.~Meyer}
\affiliation{Department of Statistics, The University of Auckland, Auckland, New Zealand}

@@

\bigskip\noindent\textit{Acknowledgments} ---

>>>>

\definecolor{maob}{rgb}{0.6,0.1,0.2}
\newcommand{\maob}[1]{\textcolor{maob}{#1}}
\newcommand{\pcd}[1]{\textcolor{blue}{#1}}
\newcommand{\mab}[1]{\textcolor{red}{#1}}
\newcommand{\tf}[1]{\textcolor{red}{TF: #1}}

@@

\begin{document}
%\linenumbers

\title{Inference of proto-neutron star properties from gravitational-wave data in core-collapse supernovae}

\author{Marie-Anne~Bizouard}
\affiliation{Artemis, Universit\'e C\^ote d'Azur, Observatoire C\^ote d'Azur, CNRS, CS 34229, F-06304 Nice Cedex 4, France}

\author{Patricio~Maturana-Russel}
\affiliation{Department of Statistics, The University of Auckland, Auckland, New Zealand}
\affiliation{Department of Mathematical Sciences, Auckland University of Technology, Auckland, New Zealand}

\author{Alejandro~Torres-Forn\'e}
\affiliation{Max Planck Institute for Gravitationalphysik (Albert Einstein Institute), D-14476 Potsdam-Golm, Germany}
\affiliation{Departamento de Astronom\'{\i }a y Astrof\'{\i }sica, Universitat de Val\`encia, E-46100 Burjassot, Val\`encia, Spain}

\author{Martin~Obergaulinger}
\affiliation{Departamento de Astronom\'{\i }a y Astrof\'{\i }sica, Universitat de Val\`encia, E-46100 Burjassot, Val\`encia, Spain}

\author{Pablo~Cerd\'a-Dur\'an}
\affiliation{Departamento de Astronom\'{\i }a y Astrof\'{\i }sica, Universitat de Val\`encia, E-46100 Burjassot, Val\`encia, Spain}

\author{Nelson~Christensen}
\affiliation{Artemis, Universit\'e C\^ote d'Azur, Observatoire C\^ote d'Azur, CNRS, CS 34229, F-06304 Nice Cedex 4, France}
\affiliation{Carleton College, Northfield, MN 55057, USA}

\author{Jos\'e~A.~Font}
\affiliation{Departamento de Astronom\'{\i }a y Astrof\'{\i }sica, Universitat de Val\`encia, E-46100 Burjassot, Val\`encia, Spain}
\affiliation{Observatori Astron\`omic, Universitat de Val\`encia, E-46980, Paterna, Val\`encia, Spain}

\author{Renate~Meyer}
\affiliation{Department of Statistics, The University of Auckland, Auckland, New Zealand}

@@

\bigskip\noindent\textit{Acknowledgments.--} Work supported by the Spanish Agencia Estatal de Investigaci\'on (PGC2018-095984-B-I00), by the Generalitat Valenciana (PROMETEO/2019/071)
and by  the European Union’s Horizon 2020 research and innovation (RISE)  programme   H2020-MSCA-RISE-2017  GrantNo.  FunFiCO-777740.  \tf{I have included my grants. Please include yours.}

### 2020-11-08 conclusion.tex
Finished my revision of the entire paper.

In our study we have used a set of 1D CCSN simulations to build a model that relates the  evolution of PNS properties with the frequency of the dominant g-mode, namely the $\mbox{}^2g_2$ mode. This relationship is extracted from the GW data using an algorithm developed for this investigation. This algorithm is  a first attempt to infer the time evolution of a particular combination of the PNS mass and radius based on the universal relations found in~\cite{Torres:2019b}. More precisely, we have considered the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ derived from the observation of the $\mbox{}^2g_2$ oscillation mode in the numerically generated GW data. The  performance of our method has been estimated employing  simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state.

We have investigated the performance of the algorithm in the case of an optimally oriented source detected in a singe GW detector. The signals of our sample have been injected into 100 Gaussian noise realisations whose PSD follow the spectra of the different GW detectors we have analyzed. For Advanced LIGO and Advanced Virgo, the ratio $r$ can be reconstructed with a good accuracy in the case of a galactic CCSN (i.e.~distances of ${\cal O}(10\, {\rm kpc})$). This holds for a wide range of progenitor masses, the quality of the inference mainly depending on the signal-to-noise ratio of the signal. For third-generation GW detectors such as the Einstein Telescope and the Cosmic Explorer, however, we obtain an order of magnitude improvement, as the $\mbox{}^2g_2$ ratio can be reconstructed for sources at distances of ${\cal O}(100\, {\rm kpc})$. In particular, Cosmic Explorer in its stage 2 configuration yields the best performance for all waveforms we have considered thanks to its excellent sensitivity in the \unit[100-1000]{Hz} range. Among the three configuration of the Einstein Telescope, ET-D provides the best performance, especially for our set waveforms with the higher progenitor masses (25 $\Msol$ and 40 $\Msol$). Comparing the estimated distances for ET-B and the other third-generations detectors, having a good sensitivity below \unit[200]{Hz} seems the most important factor to detect high mass progenitor signals.

In the present study we have assumed that the sources are optimally oriented. The reported distances at which
we can infer the time evolution of $M_{\rm PNS}/R_{\rm PNS}^2$ must thus be regarded as upper limits. Those figures may decrease by a factor 2--3 on average for a source located anywhere in the sky. Furthermore, we have not considered the detectability prospects of  CCSN waveforms in the realistic case in which the  interferometers operated as a detector network. 
We defer an improved implementation of our approach for a forthcoming publication. Finally, we note that the method discussed in this work can be adapted to other PNS oscillation modes, by simply changing a few parameters such as the initial frequency range of the mode and its monotonic raise or descent. Being able to reconstruct several modes in the same GW signal would potentially allow to individually infer the mass and the radius of the PNS.

>>>>

In our study we have used a set of 1D CCSN simulations to build a model that relates the  evolution of PNS properties with the frequency of the dominant g-mode, namely the $\mbox{}^2g_2$ mode. This relationship is extracted from the GW data using an algorithm developed for this investigation. This algorithm is  a first attempt to infer the time evolution of a particular combination of the PNS mass and radius based on the universal relations found in~\cite{Torres:2019b}. More precisely, we have considered the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ (the PNS surface gravity) derived from the observation of the $\mbox{}^2g_2$ oscillation mode in the numerically generated GW data. The  performance of our method has been estimated employing  simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state.

We have investigated the performance of the algorithm in the case of an optimally oriented source detected by a singe GW detector. Our numerical signals have been injected into 100 Gaussian noise realisations whose PSD follow the spectra of the different GW detectors analyzed. We have found that for Advanced LIGO and Advanced Virgo, the ratio $r$ can be reconstructed with a good accuracy in the case of a galactic CCSN (i.e.~for distances of ${\cal O}(10\, {\rm kpc})$). This holds for a wide range of progenitor masses, the quality of the inference mainly depending on the signal-to-noise ratio of the event. For third-generation GW detectors such as the Einstein Telescope and the Cosmic Explorer, however, we obtain an order of magnitude improvement, as the $\mbox{}^2g_2$ ratio can be reconstructed for sources at distances of ${\cal O}(100\, {\rm kpc})$. In particular, Cosmic Explorer in its stage 2 configuration yields the best performance for all waveforms we have considered thanks to its excellent sensitivity in the \unit[100-1000]{Hz} range. Among the three configurations of the Einstein Telescope, ET-D provides the best performance, especially for our set waveforms with the highest progenitor masses (25 $\Msol$ and 40 $\Msol$). Comparing the estimated distances for ET-B and the other third-generations detectors, having a good sensitivity below \unit[200]{Hz} seems the most important factor to detect high mass progenitor signals.

In the present study we have assumed that the sources are optimally oriented. The reported distances at which
we can infer the time evolution of $M_{\rm PNS}/R_{\rm PNS}^2$ must thus be regarded as upper limits. Those figures may decrease by a factor 2--3 on average for a source located anywhere in the sky. Furthermore, we have not considered the detectability prospects of  CCSN waveforms in the realistic case in which the  interferometers operated as a detector network. 
We defer an improved implementation of our approach for a forthcoming publication. Finally, we note that the method discussed in this work can be adapted to other PNS oscillation modes, by simply changing a few parameters such as the initial frequency range of the mode and its monotonic raise or descent. Being able to reconstruct several modes in the same GW signal would potentially allow to individually infer the mass and the radius of the PNS in core-collapse supernova explosions.

### 2020-11-08 introduction.tex
Finished my revision of the entire paper.

In the last decade significant progress has been made in the development of numerical codes, {in particular in the treatment of multidimensioal effects \citep{BMueller:2020}.} In the case of  neutrino-driven explosions, the GW emission is {primarly induced by instabilities developed at the newly formed proto-neutron star (PNS) and by the non-spherical accreting flow of hot matter over its surface \citep{Kotake:2017}.  This} dynamics excite the different modes of oscillation of the PNS which ultimately leads to the emission of GWs. The frequency and time evolution of these modes carry information about the properties of the GW emitter and could allow to perform PNS asteroseismology.

@@

All multidimensional numerical simulations show the systematic appearance in time-frequency diagrams (or spectrograms) of a distinct and relatively narrow feature 
%
%{The main feature appearing systematically in the GW spectrum of multidimensional numerical simulations is a strong and relatively narrow \mab{oscillation} 
%
during the post-bounce evolution of the system, with raising frequency
from about $100$~Hz up to a few kHz (at most) and a typical duration of $0.5-1$~s. This feature has been interpreted as a continuously excited gravity mode (g-mode, see \citep{kokkotas,Friedman:2013} for a definition in this context) of the PNS \citep{murphy:09, mueller:13gw, Cerda:2013, Yakunin:2015, Kuroda:2016, Andresen:2017}. 
In these models the monotonic raise of the frequency of the mode is related to the contraction of the PNS. The {typical} frequencies of {these} modes make them interesting targets for
%promising {source} for 
ground-based GW interferometers.

>>>>

In the last decade significant progress has been made in the development of numerical codes, {in particular in the treatment of multidimensioal effects \citep{BMueller:2020}.} In the case of  neutrino-driven explosions, the GW emission is {primarly induced by instabilities developed at the newly formed proto-neutron star (PNS) and by the non-spherical accreting flow of hot matter over its surface \citep{Kotake:2017}.  These} dynamics excite the different modes of oscillation of the PNS which ultimately leads to the emission of GWs. The frequency and time evolution of these modes carry information about the properties of the GW emitter and could allow to perform PNS asteroseismology.

@@

All multidimensional numerical simulations show the systematic appearance in time-frequency diagrams (or spectrograms) of a distinct and relatively narrow feature 
%
%{The main feature appearing systematically in the GW spectrum of multidimensional numerical simulations is a strong and relatively narrow \mab{oscillation} 
%
during the post-bounce evolution of the system, with frequency rising 
from about $100$~Hz up to a few kHz (at most) and a typical duration of $0.5-1$~s. This feature has been interpreted as a continuously excited gravity mode (g-mode, see \citep{kokkotas,Friedman:2013} for a definition in this context) of the PNS \citep{murphy:09, mueller:13gw, Cerda:2013, Yakunin:2015, Kuroda:2016, Andresen:2017}. 
In these models the monotonic raise of the frequency of the mode is related to the contraction of the PNS. The {typical} frequencies of {these} modes make them interesting targets for
%promising {source} for 
ground-based GW interferometers.

### 2020-11-08 methods.tex
Finished my revision of the entire paper.

The best fitting model amongst polynomials of degree 1, 2, and 3  was chosen according to
the Aikaike information criterion with coefficients given in Table \ref{tab:model}, which is actually the model defined in \eqref{eq:model1}.  The data and fit of the model including 95\% confidence bands are displayed in
Figure~\ref{fig:LMVAR}.

@@

{We use this model to infer the properties of the simulations in the 
{\it test set} discussed in Section \ref{sec:simulations}.}
To describe the method we focus on the GW signal
of model {\texttt s20S}, originally
sampled at 16384 Hz but downsampled at 4096 Hz.
A spectrogram of this signal is shown in Figure \ref{fig:spectrogram} based on
autoregressive estimates \pcd{[add ref?]} of the local spectra for successive time intervals of 
length 200 with a 90\% overlap.
The dominant emission mode corresponds to the PNS oscillation $\mbox{}^2 g_2$-mode. We have
developed a time-frequency method to track the ridge $m(t) $ in the spectrogram,
taking into account that it is monotonically increasing with time. 
This is a property of the $\mbox{}^2 g_2$-mode whose frequency   
increases as the PNS becomes more massive and compact.
Starting from either the left- or right-most column of the time-frequency matrix
we identify and trace the sequence of amplitude peaks within a certain frequency
band given the monotonicity constraint. Specific details about the reconstruction of the $\mbox{}^2 g_2$-mode ridge 
are provided in Appendix \ref{app:gmode}.

>>>>

The best fitting model amongst polynomials of degree 1, 2, and 3  was chosen according to
the Akaike information criterion with coefficients given in Table \ref{tab:model}, which is actually the model defined in \eqref{eq:model1}.  The data and fit of the model including 95\% confidence bands are displayed in
Figure~\ref{fig:LMVAR}.

@@

{We use this model to infer the properties of the simulations in the 
{\it test set} discussed in Section \ref{sec:simulations}.}
To describe the method we focus on the GW signal
of model {\texttt s20S}, originally
sampled at 16384 Hz but downsampled at 4096 Hz.
A spectrogram of this signal is shown in Figure \ref{fig:spectrogram} based on
autoregressive estimates~\cite{BrockwellPeterJ1991Tsta} of the local spectra for successive time intervals of 
length 200 with a 90\% overlap.
The dominant emission mode corresponds to the PNS oscillation $\mbox{}^2 g_2$-mode. We have
developed a time-frequency method to track the ridge $m(t) $ in the spectrogram,
taking into account that it is monotonically increasing with time. 
This is a property of the $\mbox{}^2 g_2$-mode whose frequency   
increases as the PNS becomes more massive and compact.
Starting from either the left- or right-most column of the time-frequency matrix
we identify and trace the sequence of amplitude peaks within a certain frequency
band given the monotonicity constraint. Specific details about the reconstruction of the $\mbox{}^2 g_2$-mode ridge 
are provided in Appendix \ref{app:gmode}.
